# **Non-Parametric Methods for Statistical Inference**

In the previous chapter we discussed several statistical tests applied to a sample from a population. The sample is assumed to be independent and identically distributed, and all values are assumed to be a random sample from a Normal distribution.

Non-parametric methods in statistics provide a versatile and robust alternative to traditional parametric techniques, particularly when the assumptions of normality or other underlying distributions cannot be met. These methods are characterized by their flexibility and ability to analyze data without making specific assumptions about the population distribution. Examples of non-parametric methods include the Wilcoxon signed-rank test, which is a non-parametric alternative to the paired t-test when assessing differences between paired observations; the Mann-Whitney U test, used for comparing two independent groups, and the Kruskal-Wallis test, which extends the comparison to multiple groups. These techniques are valuable tools for researchers and analysts working with real-world data, as they offer reliable solutions in scenarios where parametric methods may fall short.

## Introduction to One-Sample Rank Tests

One common scenario where rank tests shine is when dealing with skewed or non-normally distributed data. In such cases, traditional parametric tests like the t-test or ANOVA may produce unreliable results. Rank tests offer a solution by focusing on the ranks or orders of observations, making them robust to distributional assumptions and potential outliers.

For example, consider a study evaluating the effectiveness of a new medication. The response variable is the level of pain relief reported by patients on a scale from 1 (no relief) to 10 (complete relief). The data collected may not follow a normal distribution, and the scale itself is ordinal rather than continuous. In this situation, a rank-based test like the Wilcoxon signed-rank test for paired data or the Mann-Whitney U test for independent samples can provide a more appropriate analysis, comparing the central tendency of pain relief between groups or before and after treatment without relying on distributional assumptions.

In essence, rank tests offer a valuable tool for statisticians and researchers when dealing with real-world data that doesn't conform to the idealized assumptions of parametric methods. They provide a robust and reliable means of drawing valid conclusions, even in challenging situations.

### Wilcoxon Signed-Rank Test

The signed rank tests uses the difference between the sample values from the proposed median $x_i-m$ , but also uses the size of the difference to calculate a rank. This tests assumes that the sample is coming from a symmetrical distribution.

**A simple example**

If the differences of the sample from the proposed median are: -1.6, 2.3, -5.1, 3.9, the sorted values without considering the signs are: -1.6, 2.3, 3.9, -5.1, with ranks 1, 2, 3 and 4. The values of the signed rank test statistic ($TS$) are the ranks for the **negative values**, this is: $TS=1+4=5$. Other options are to use the positive value ranks: $TS=2+3$.

- A large value of TS would indicate more observations **below the median** than expected under $H_0$.
- When ties are present in the data, they are removed and the sample size is adjusted accordingly.
- When $n$ is moderately large, the TS statistic has an Normal distribution under the Null hypothesis, with mean and variance: $$E[TS]=\frac{n(n+1)}{4}$$ $$Var[TS]=\frac{n(n+1)(2n+1)}{24}$$

Assume that $TS=t$. The Signed Rank Test rejects the Null hypothesis for a two-tailed test if:

- $P(TS\leq t) \leq \alpha/2$
- $P(TS\geq t) \leq \alpha/2$

In R, the function *wilcox.test(.)* can be used for testing the location parameter for a single population, using the parameter $\mu=\ldots$ in the function call.

**Important**: *The sign test does not need any assumption about the shape of the distribution of the data*. The Wilcoxon Signed Rank test requires that the data has a symmetric distribution.

**Example**

In this example, we want to compare the mean duration and mean intensity of historical heat waves in the city of Champaign, Illinois before the year 2005 and after 2005.

The data is available in the file `hw_factors.csv`.

```{r, message=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(descstatsr)
library(goftest)
library(BSDA)
hwf<-read.csv("data/hw_factors.csv")
# Mean Intensity
hwf$period<- ifelse(hwf$year<2005,"Before 2005","After 2005")
hwf %>% filter(year<2005) %>% 
  summarize(mean_int_b2005=mean(mean_intensity))
hwf %>% filter(year>=2005) %>% 
  summarize(mean_int_a2005=mean(mean_intensity))

# Mean duration
hwf %>% filter(year<2005) %>% 
  summarize(mean_dur_b2005=mean(mean_duration))
hwf %>% filter(year>=2005) %>% 
  summarize(mean_dur_a2005=mean(mean_duration))

hwf1<-hwf %>% filter(year<2005)
hwf2<-hwf %>% filter (year >=2005)

# One-sample t-test for mean intensity
t.test(hwf1$mean_intensity,mu=5,alternative="l")
t.test(hwf2$mean_intensity,mu=5, alternative="l")
# Two-sample t-test for mean intensity
# By default variances are not assumed equal
t.test(hwf1$mean_intensity,hwf2$mean_intensity)

# Sign-test (One-sample) for mean duration
SIGN.test(hwf1$mean_duration, md = 3.5, alternative='two.sided')

SIGN.test(hwf2$mean_duration, md = 3.5, alternative='two.sided')

# Wilcoxon Signed-Rank test (One-sample) for mean duration

#Before 2005
wilcox.test(hwf1$mean_duration,mu=3.5)

#After 2005
wilcox.test(hwf2$mean_duration,mu=3.5)
```

## Rank Tests with Two or More Samples

Assume we have data on precipitation projections from two Global Circulation Models (GCM1 and GCM2) for the period 2070- 2100 (last 30 years of this century), and we want to test the hypothesis that the location parameters for the two distributions ($\tilde{\mu}_1\;\;\mbox{and}\;\;\tilde{\mu}_2$) are the same. We assume that the two samples from models GCM1 and GCM2 are not independent samples, since the models essentially have similar differential equations to represent the climate-land-ocean interactions.

We can calculate the differences between each pair of observations for every time step $x_i -y_i$, and apply the Wilcoxon Rank test or the Sign test.

For a comparison of the location parameters among 3 or more populations we can use the `Kruskal-Wallis` test. The parametric parallel of this test is the one-way ANOVA model that will be discussed later.

### Wilcoxon Rank-Sum test

The `Wilcoxon Rank Sum Test`, also known as the Mann-Whitney U test, is a non-parametric test used to assess whether there is a difference between two independent groups. It doesn't assume normality and it is appropriate when the data may not meet the assumptions of a t-test.

We want to show that observations from one population $x_1, x_2,\ldots, x_{n_1}$ tend to be larger (or smaller) than observations from another population $y_1, y_2, \ldots, y_{n_2}$.

It is assumed that the distribution $F_1$ of $X$ and $F_2$ of $Y$ are identical in shape and only differ by location. This is equivalent to say that there is a shift in location.

In this case we set up the hypothesis (more general set up): $$H_0: F_1=F_2\;\;\mbox{vs.}\;\; H_a: F_1> F_2$$ (one-sided) or $$H_0: F_1=F_2\;\; \mbox{vs.}\;\; H_a: F_1 > F_2\;\;\mbox{or} \;\; H_a: F_1 < F_2 $$ (two-sided).

If the data support the alternative hypothesis $H_a: F_1< F_2$ we can say that random variable $Y$ has **stochastic dominance** over random variable $X$.

The steps for the test are as follows (assume no ties are present): 1. Rank all $N=n_1+n_2$ observations in ascending order; 2. Sum the ranks of the $x's$ ($W_1=\omega_1$) and the ranks of the $y's$ ($W_2=\omega_2$); 3. Reject $H_0$ if $\omega_1$ is large and equivalently $\omega_2$ is small.

The distributions of $W_1$ and $W_2$ are not the same when the sample sizes $n_1$ and $n_2$ are not equal.

### Wilcoxon rank sum test and Mann-Whitney test

The **Mann-Whitney U test** is another non-parametric test available when you want to compare two distributions. The probability distribution of the random variable $W$ (sum of the ranks) is tabulated. *The Mann-Whitney test* statistic $U$ is related to the Wilcoxon rank sum test. In the Mann-Whitney test, the steps are as follows:

1.  Calculate $u_1$= number of pairs in which $x_i>y_j$ for all possible pairs $i$ and $j$; and $u_2$= number of pairs in which $x_i<y_j$ for all possible pairs $i$ and $j$. There are $n_1n_2$ possible pairs and $u_1+u_2=n_1n_2$.
2.  Reject $H_0$ if $u_1$ is large or $u_2$ is small.

The Wilcoxon rank sum test and Mann-Whitney test are related in the following way: $$u_1=\omega_1 - \frac{n_1(n_1+1)}{2}\;\;\mbox{and}\;\;u_2=\omega_2 - \frac{n_2(n_2+1)}{2}$$

The upper tail probability distributions for the $U$ and $W$ statistics under $H_0$ are tabulated. An $\alpha$-level test rejects $H_0$ if $P-value\leq \alpha$

**Example**

Compare mean heat wave's duration before and after 2005

We want to check whether there has been a shift in the distributions of the mean heat wave's duration after 2005.

We will use the function *wilcox.test(x,y)*, where $x$ is the sample vector from the random variable $X$, and $y$ is the sample vector from random variable $Y$.

```{r}
wilcox.test(hwf1$mean_duration,hwf2$mean_duration)
```

For a comparison of the location parameters among 3 or more populations we can use the Kruskal-Wallis test.

### Kruskal-Wallis test

The Kruskal-Wallis test is a powerful non-parametric statistical test designed to address situations where the assumptions of normality and homogeneity of variances required by traditional analysis of variance (ANOVA) cannot be met. This test is particularly useful when you have three or more independent groups or conditions and you want to determine whether there are statistically significant differences in the central tendencies of these groups.

In cases where the data is not normally distributed or the group variances are unequal, the Kruskal-Wallis test provides a reliable alternative to traditional ANOVA. Instead of relying on means and variances, the Kruskal-Wallis test operates on the ranks of the data, making it robust to deviations from normality and homoscedasticity.

Imagine a clinical trial comparing the pain relief efficacy of three different treatments. Patients' pain relief scores may not follow a normal distribution, and the variability in responses between treatments could be quite different. In this scenario, the Kruskal-Wallis test allows researchers to determine whether there are statistically significant differences in pain relief among the treatments without the need for stringent parametric assumptions.

In summary, the Kruskal-Wallis test is a valuable tool in situations where you need to compare multiple groups and the underlying data does not meet the assumptions of ANOVA. It provides a reliable means of detecting differences in central tendencies across groups while accommodating the realities of real-world data. When the number of groups is equal to two, the Kruskall-Wallis test reduces to the Wilcoxon Rank Sum test.

**Example**

Increased greenhouse gas concentrations in the atmosphere as a consequence of human activities, might modify the behaviour of rainfall extreme events. To investigate that, we use data from the CMIP5 project consisting of simulated values of daily rainfall for three different scenarios: historical conditions, RCP4.5 and RCP8.5 climate scenarios proposed by the IPCC, being the last one the worst case scenario with the highest radiative forcing level (8.5 $Watts/m^2$) for the projected future climate. We want to study potential changes in the location parameter of the maximum annual daily rainfall. The Generalized Extreme Value (GEV) distribution (to be discussed in Chapter XX) was fitted to the simulated daily annual maximum for the three experiments (historical, RCP4.5 and RCP8.5) using simulated data from 30 General Circulation Models (GCMs). The location, scale and shape parameters of the GEV fitted models were estimated using maximum likelihood; and the Kruskal-Wallis test was used to assess whether significant differences in the model parameters for the three experiments are observed.

## Non-parametric correlation measures

### The Spearman's rank correlation

The Spearman's correlation ($\rho_S$) measures the strength of a monotonic relationship between two variables. When the relationship between two variables is non-linear but there is a monotonic relationship between then, the Spearman's correlation coefficient is a better measure of the association between the two variables. This a non-parametric measure of correlation and it is calculated as the Pearson correlation coefficient between the ranks of the two variables with sample estimate:

$$r_{s_{xy}}= \frac{cov(Rank(x),Rank(y))}{S_{Rank(x)}\times S_{Rank(y)}}$$

where `cov` is the covariance between the ranks of the two variables and $S_{rank(.)}$ is the standard deviation of the ranks for each variable. This formula is equivalent to the Pearson correlation formula between the ranks of the two samples.

The significance of $\rho_S$ can be tested by using the R function `cor.test` by setting the function parameter `method=spearman.`

### Kendal's tau correlation

It is more robust and efficient than the Spearman's correlation, but more difficult to interpret: Considering all possible pairs $(i,j)$ from the sample. The Kendall's $\tau$ quantifies the difference between the percentage of all discordant (ranks disagree) and concordant pairs (ranks agree) in the data: $$\tau_{xy}=\frac{2}{n(n-1)}\sum_{i<j} sign(x_i-x_j)\times sign(y_i-y_j)$$

Kendall's $\tau$ is recommended for small sample size or when there are many ties in the data. A significance test is also implemented using the `cor.test` function with `method=kendall`.

When there relationship between two variables is highly non-linear or there is presence of outliers, the Spearman and Kendall's $\tau$ are recommended over the Pearson correlation coefficient.

## Bootstrap standard errors and  Confidence Intervals

Another class of non-parametric methods, known as re-sampling methods, are based on drawing random samples from the given sample to make inferences about the population. An example of this is the bootstrap method, which is useful to calculate uncertainty estimates of the sample paremeters and confidence intervals of the true parameter values. Usually these uncertainty estimates are measured through the standard errors.

### Bootstrapping

**Example**
